---
title: "Alzheimer's disease PGS, APOE status, and summary cognitive status in the HRS"
subtitle: "Inclusion/Exclusion, Flowchart & Univariate Table"
author: "Kelly Bakulski"
date: "October 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Scripts/")
```

## Goal: Producing reproducible inclusion/exclusion descriptive statistics tables, flowcharts

```{r set source/output locations}
#Setting file paths for data, output, scripts

source_dir<-"~/Google Drive/AD_R01_WB/Data/" #location of the parent dataset

results_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Results/" #location of output tables
  
derived_data_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Derived_Datasets/" #location of any new datasets I'm building specifically for this paper

script_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Scripts/" #location of this script, redundant with the knitr option above

date<-format(Sys.Date(), "%Y%m%d")
```

#Load in the dataset from Rda: HRS 1995-2014
```{r load}
load(file = paste0(derived_data_dir,"hrs_democoggen_wide20190813.rda"))
dim(hrs)
```
# Parent dataset file "~/Google Drive/AD_R01_WB/Data/hrs_democoggen_wide20190805.sas7bdat"
# Reference file "~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Scripts/HRS_PGS_AD_Cog_DataQC_kmb_20190809.Rmd" for derived variable development


## Need ancestry agnostic genetic variables, dummy missing variables for univariate table
```{r missing variable update}


#PH13_KunkleOK2_PT_0.01
#PH13_KunkleFULL_PT_0.01

### Secondary PGS variable, with APOE
summary(hrs$EAPH13_KunkleFULL_PT_0.01)
summary(hrs$AAPH13_KunkleFULL_PT_0.01)
table(is.na(hrs$EAPH13_KunkleFULL_PT_0.01), is.na(hrs$AAPH13_KunkleFULL_PT_0.01)) # Expect new variable to have 30962 missing

hrs$PH13_KunkleOK2_PT_0.01.miss <- ifelse(is.na(hrs$PH13_KunkleOK2_PT_0.01), "(Missing)", NA)
table(hrs$PH13_KunkleOK2_PT_0.01.miss)

hrs$PH13_KunkleFULL_PT_0.01.miss <- ifelse(is.na(hrs$PH13_KunkleFULL_PT_0.01), "(Missing)", NA)
table(hrs$PH13_KunkleFULL_PT_0.01.miss)

### Main APOE variable
table(hrs$APOE2010_bin, useNA = "always")
hrs$APOE2010_bin.fac <- factor(hrs$APOE2010_bin, levels = c(0, 1), labels = c("None", "E4"))
# hrs$APOE2010_bin.fac<-ifelse(is.na(hrs$APOE2010_bin.fac), "Missing", hrs$APOE2010_bin.fac)
# hrs$APOE2010_bin.fac<-factor(hrs$APOE2010_bin.fac, levels=c(1,2,3), labels=c("None", "E4", "Missing"))
table(hrs$APOE2010_bin.fac, useNA = "always")
hrs$APOE2010_bin.miss <- ifelse(is.na(hrs$APOE2010_bin), "(Missing)", NA)
table(hrs$APOE2010_bin.miss)

### Secondary APOE multiple level variable
# Okay as APOE2010_imputed
table(hrs$APOE2010_imputed, useNA = "always")
hrs$APOE2010_imputed.miss <- ifelse(is.na(hrs$APOE2010_imputed), "(Missing)", NA)
table(hrs$APOE2010_bin.miss)

# Need outcome varaiable as a factor
table(hrs$DC_60_2V, useNA = "always")
hrs$DC_60_2V.fac <- factor(hrs$DC_60_2V, levels = c(1:5), labels = c("Dementia", "CIND", "Borderline", "Normal", "Unclassified"))
table(hrs$DC_60_2V.fac, useNA = "always")
hrs$DC_60_2V.miss <- ifelse(is.na(hrs$DC_60_2V), "(Missing)", NA)
table(hrs$DC_60_2V.miss, useNA = "always")

hrs$DC_60_2V.fac.relevel <- factor(hrs$DC_60_2V.fac, levels = c("Normal", "Borderline", "CIND", "Dementia"), labels = c("Normal", "Borderline", "Impaired", "Dementia"))
table(hrs$DC_60_2V.fac.relevel)

hrs$DC_60_2V.relevel<-ifelse(hrs$DC_60_2V==4, 1, NA)
hrs$DC_60_2V.relevel<-ifelse(hrs$DC_60_2V==3, 2, hrs$DC_60_2V.relevel)
hrs$DC_60_2V.relevel<-ifelse(hrs$DC_60_2V==2, 3, hrs$DC_60_2V.relevel)
hrs$DC_60_2V.relevel<-ifelse(hrs$DC_60_2V==1, 4, hrs$DC_60_2V.relevel)
table(hrs$DC_60_2V.relevel)

## Need sex as a factor
table(hrs$SEX, useNA = "always")
hrs$SEX.fac <- factor(hrs$SEX, levels = c(1, 2), labels = c("Male", "Female"))
table(hrs$SEX.fac, useNA = "always")
hrs$SEX.miss <- ifelse(is.na(hrs$SEX.fac), "(Missing)", NA)
table(hrs$SEX.miss, useNA = "always")

### Missing dummy variable for education
summary(hrs$SCHLYRS)
hrs$SCHLYRS.miss <- ifelse(is.na(hrs$SCHLYRS), "(Missing)", NA)
table(hrs$SCHLYRS.miss)

summary(hrs$AGE_LAST_60)
hrs$AGE_LAST_60.miss <- ifelse(is.na(hrs$AGE_LAST_60), "(Missing)", NA)
table(hrs$AGE_LAST_60.miss)

summary(hrs$YEAR_LAST_60)
hrs$YEAR_LAST_60.miss <- ifelse(is.na(hrs$YEAR_LAST_60), "(Missing)", NA)
table(hrs$YEAR_LAST_60.miss)

table(hrs$ancestry, useNA = "always")
hrs$ancestry.miss <- ifelse(is.na(hrs$ancestry), "(Missing)", NA)
table(hrs$ancestry.miss)

hrs$ancestry.relevel <- factor(hrs$ancestry, levels = c("European", "African"))
table(hrs$ancestry.miss)

table(hrs$AHEAD_CODA, useNA = "always")
hrs$AHEAD_CODA.miss <- ifelse(is.na(hrs$AHEAD_CODA), "(Missing)", NA)
table(hrs$AHEAD_CODA.miss)

summary(hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2.miss <- ifelse(is.na(hrs$BMI_LAST_60_v2), "(Missing)", NA)
table(hrs$BMI_LAST_60_v2.miss)

table(hrs$HBP_EVER_LAST_60, useNA = "always")
hrs$HBP_EVER_LAST_60.miss <- ifelse(is.na(hrs$HBP_EVER_LAST_60), "(Missing)", NA)
table(hrs$HBP_EVER_LAST_60.miss)

table(hrs$T2D_EVER_LAST_60, useNA = "always")
hrs$T2D_EVER_LAST_60.miss <- ifelse(is.na(hrs$T2D_EVER_LAST_60), "(Missing)", NA)
table(hrs$T2D_EVER_LAST_60.miss)

table(hrs$EVSTROKE_LAST_60, useNA = "always")
hrs$EVSTROKE_LAST_60.miss <- ifelse(is.na(hrs$EVSTROKE_LAST_60), "(Missing)", NA)
table(hrs$EVSTROKE_LAST_60.miss)

table(hrs$CESD_BIN, useNA = "always")
hrs$CESD_BIN.miss <- ifelse(is.na(hrs$CESD_BIN), "(Missing)", NA)
table(hrs$CESD_BIN.miss)

table(hrs$SMK_CURR_LAST_60, useNA = "always")
hrs$SMK_CURR_LAST_60.miss <- ifelse(is.na(hrs$SMK_CURR_LAST_60), "(Missing)", NA)
#table(hrs$SMK_CURR_LAST_60.miss, hrs$include.step1)

table(hrs$ALC_CURR_LAST_60, useNA = "always")
hrs$ALC_CURR_LAST_60.miss <- ifelse(is.na(hrs$ALC_CURR_LAST_60), "(Missing)", NA)
table(hrs$ALC_CURR_LAST_60.miss)

# library(forcats)
# table(fct_explicit_na(hrs$DC_60_2V.fac, na_level = "(Missing)"))
```


Make exclusions (less than 3 visits, younger than 60, missing APOE, missing SCHLYRS, health behaviors), confirm matching SAS exclusions
```{r exclusions}
#Start with the 2010 PGS _T

hrs$unique <- paste0(hrs$HHID, hrs$PN)
# Step 1: Exclude missing genetic information 2010 polygenic scores
#PH13_KunkleYAPOE_PT_1
#PH13_KunkleOK2_PT_0.01

table(!is.na(hrs$EAPH13_KunkleOK2_PT_0.01 | hrs$AAPH13_KunkleOK2_PT_0.01)) # Step 1 of flow chart

step1 <- hrs[!is.na(hrs$PH13_KunkleOK2_PT_0.01), ]
dim(step1) # checks out
hrs$include.step1 <- ifelse(!is.na(hrs$PH13_KunkleOK2_PT_0.01), 1, 0)
table(hrs$include.step1)
# Note: Previous tables started at step 2 (those excluded at step 1 were not in the inclusion/exclusion table)

# Step 2: At least 3 cognition variables
table(!is.na(hrs$DC_60_2V), hrs$include.step1) # Differs from flow chart in expected direction, given we're including stroke
step2 <- step1[!is.na(step1$DC_60_2V), ]
dim(step2) 
hrs$include.step2 <- ifelse(hrs$include.step1 == 1 & !is.na(hrs$DC_60_2V), 1, 0)
table(hrs$include.step2)

# Step 3: Drop unclassified cognition, think this can come out in the previous step
#table(step2$DC_60_2V) 
step3 <- step2[step2$DC_60_2V < 5, ]
dim(step3) 
table(hrs$include.step2, hrs$DC_60_2V==5)
hrs$include.step3 <- ifelse(hrs$include.step2 == 1 & hrs$DC_60_2V < 5, 1, 0)
table(hrs$include.step3)

# Step 4: Drop missing SCHLYRS
step4 <- step3[!is.na(step3$SCHLYRS), ]
dim(step4)
table(hrs$include.step3, !is.na(hrs$SCHLYRS), !is.na(hrs$BMI_LAST_60_v2))
hrs$include.step4 <- ifelse(hrs$include.step3 == 1 & !is.na(hrs$SCHLYRS), 1, 0)

table(hrs$include.step4, hrs$include.step3) #no missing on SCHLYRS
table(hrs$include.step4, !is.na(hrs$BMI_LAST_60_v2))
table(hrs$include.step4, hrs$include.step3) # Dropped 11

# Step 5: Drop missing BMI
step5 <- step4[!is.na(step4$BMI_LAST_60_v2), ]
dim(step5)
table(hrs$include.step4, !is.na(hrs$BMI_LAST_60_v2))
hrs$include.step5 <- ifelse(hrs$include.step4 == 1 & !is.na(hrs$BMI_LAST_60_v2), 1, 0)
table(hrs$include.step4)
summary(step5$BMI_LAST_60_v2)
summary(hrs$BMI_LAST_60_v2[hrs$include.step5==1])
table(hrs$include.step5)
table(hrs$include.step5, hrs$include.step4) #Dropped 88

# Step 6: Drop missing Smk
table(hrs$include.step5, !is.na(hrs$SMK_CURR_LAST_60))
step6 <- step5[!is.na(step5$SMK_CURR_LAST_60), ]
dim(step6)
hrs$include.step6 <- ifelse(hrs$include.step5 == 1 & !is.na(hrs$SMK_CURR_LAST_60), 1, 0)
table(hrs$include.step6)
table(step6$SMK_CURR_LAST_60)
table(hrs$SMK_CURR_LAST_60[hrs$include.step5==1])
table(hrs$include.step6, hrs$include.step5) #dropped 64

# Step 7: Drop missing HBP
table(hrs$include.step6, !is.na(hrs$HBP_EVER_LAST_60)) #expect 25
step7 <- step6[!is.na(step6$HBP_EVER_LAST_60), ]
dim(step7)
hrs$include.step7 <- ifelse(hrs$include.step6 == 1 & !is.na(hrs$HBP_EVER_LAST_60), 1, 0)
table(hrs$include.step7)
table(step7$HBP_EVER_LAST_60)
table(hrs$HBP_EVER_LAST_60[hrs$include.step6==1])
table(hrs$include.step7, hrs$include.step6) #dropped 25

# Step 8: Drop missing T2D
table(hrs$include.step7, !is.na(hrs$T2D_EVER_LAST_60)) #expect 14
step8 <- step7[!is.na(step7$T2D_EVER_LAST_60), ]
dim(step8)
hrs$include.step8 <- ifelse(hrs$include.step7 == 1 & !is.na(hrs$T2D_EVER_LAST_60), 1, 0)
table(hrs$include.step8)
table(step8$T2D_EVER_LAST_60)
table(hrs$T2D_EVER_LAST_60[hrs$include.step8==1])
table(hrs$include.step8, hrs$include.step7) #dropped 14

# Step 9: Drop missing stroke
table(hrs$include.step8, !is.na(hrs$EVSTROKE_LAST_60)) #expect 12
step9 <- step8[!is.na(step8$EVSTROKE_LAST_60), ]
dim(step9)
hrs$include.step9 <- ifelse(hrs$include.step8 == 1 & !is.na(hrs$EVSTROKE_LAST_60), 1, 0)
table(hrs$include.step8)
table(step9$EVSTROKE_LAST_60)
table(hrs$EVSTROKE_LAST_60[hrs$include.step9==1])
table(hrs$include.step9, hrs$include.step8) #dropped 12

# Step 10: Drop missing alcohol
table(hrs$include.step9, !is.na(hrs$ALC_CURR_LAST_60)) #expect 10
step10 <- step9[!is.na(step9$ALC_CURR_LAST_60), ]
dim(step10)
hrs$include.step10 <- ifelse(hrs$include.step9 == 1 & !is.na(hrs$ALC_CURR_LAST_60), 1, 0)
table(hrs$include.step10)
table(step10$ALC_CURR_LAST_60)
table(hrs$ALC_CURR_LAST_60[hrs$include.step10==1])
table(hrs$include.step10, hrs$include.step9) #dropped 10

# Step 11: Drop missing depression
table(hrs$include.step10, !is.na(hrs$CESD_BIN)) #expect 0
# step11 <- step10[!is.na(step10$CESD_BIN), ]
# dim(step11)
# hrs$include.step11 <- ifelse(hrs$include.step10 == 1 & !is.na(hrs$CESD_BIN), 1, 0)
# table(hrs$include.step11)
# table(step11$CESD_BIN)
# table(hrs$CESD_BIN[hrs$include.step11==1])
# table(hrs$include.step11, hrs$include.step10) #dropped 0

#BMI_LAST_60_v2, EVSTROKE_LAST_60
### Use new variables: SMK_CURR_LAST_60, ALC_CURR_LAST_60, HBP_EVER_LAST_60, T2D_EVER_LAST_60, CESD_BIN

# table(!is.na(hrs$BMI_LAST_60_v2) & !is.na(hrs$EVSTROKE_LAST_60) & !is.na(hrs$SMK_CURR_LAST_60) & !is.na(hrs$ALC_CURR_LAST_60) & !is.na(hrs$HBP_EVER_LAST_60) & !is.na(hrs$T2D_EVER_LAST_60) & !is.na(hrs$CESD_BIN),hrs$include.step5,  useNA = "always")
# 
# dim(hrs[!is.na(hrs$BMI_LAST_60_v2) & !is.na(hrs$EVSTROKE_LAST_60) & !is.na(hrs$SMK_CURR_LAST_60) & !is.na(hrs$ALC_CURR_LAST_60) & !is.na(hrs$HBP_EVER_LAST_60) & !is.na(hrs$T2D_EVER_LAST_60) & !is.na(hrs$CESD_BIN) & hrs$include.step5==1,])
# 
# step6 <- hrs[!is.na(hrs$BMI_LAST_60_v2) & !is.na(hrs$EVSTROKE_LAST_60) & !is.na(hrs$SMK_CURR_LAST_60) & !is.na(hrs$ALC_CURR_LAST_60) & !is.na(hrs$HBP_EVER_LAST_60) & !is.na(hrs$T2D_EVER_LAST_60) & !is.na(hrs$CESD_BIN) & hrs$include.step5==1, ] 
# dim(step6)
length(hrs$unique[step10$unique])

# hrs$include.step6 <- ifelse(!is.na(hrs$BMI_LAST_60_v2) & !is.na(hrs$EVSTROKE_LAST_60) & !is.na(hrs$SMK_CURR_LAST_60) & !is.na(hrs$ALC_CURR_LAST_60) & !is.na(hrs$HBP_EVER_LAST_60) & !is.na(hrs$T2D_EVER_LAST_60) & !is.na(hrs$CESD_BIN) & hrs$include.step5==1, 1, 0)
# table(hrs$include.step6)


# Summary
#table(hrs$unique %in% step4$unique)
table(hrs$unique %in% hrs$unique[hrs$include.step10==1])

hrs$include <- ifelse(hrs$unique %in% hrs$unique[hrs$include.step10==1], "Included", "Excluded")

include<-hrs[hrs$include=="Included",]
dim(include)
table(hrs$include, hrs$include.step10)
table(hrs$ancestry, hrs$include)
```
### Made a new variable: include for if person made it in the primary analytic sample

### Figure out which health behaviors are missing
```{r , eval=F}
install.packages("expss")
library(expss)
table(hrs$include.step4, hrs$include.step6)
test<-hrs[hrs$include.step6==0 & hrs$include.step5==1,]
cro(is.na(test$EVSTROKE_LAST_60), is.na(test$BMI_LAST_60_v2), is.na(test$SMK_CURR_LAST_60)) #88 additional BMI, 11 schoolyears, 14 stroke, 1 smk

table(is.na(test$BMI_LAST_60_v2)) #88
table(is.na(test$SMK_CURR_LAST_60)) #65
table(is.na(test$HBP_EVER_LAST_60)) #26
table(is.na(test$T2D_EVER_LAST_60)) #17
table(is.na(test$EVSTROKE_LAST_60)) #14
table(is.na(test$ALC_CURR_LAST_60)) #11


table(!is.na(hrs$BMI_LAST_60_v2) & !is.na(hrs$EVSTROKE_LAST_60) & !is.na(hrs$SMK_CURR_LAST_60) & !is.na(hrs$ALC_CURR_LAST_60) & !is.na(hrs$HBP_EVER_LAST_60) & !is.na(hrs$T2D_EVER_LAST_60) & !is.na(hrs$CESD_BIN),hrs$include.step5,  useNA = "always")
```



### Make flow chart for inclusion/exclusion
```{r primary flowchart}
# install.packages("DiagrammeR")
library(DiagrammeR)
# install.packages("DiagrammeRsvg")
library(DiagrammeRsvg)
# install.packages("rsvg")
library(rsvg)

# graph: this sets the overall layout, themes of the plot
# node: the boxes in the flowchart
# edges: the lines used to connect each of the boxes

## Need 2 columns (1, 2) with 5 rows (A-E)
# Labels for column 1
a1 <- paste0("Polygenic scores
available
n=", length(hrs$include.step1[hrs$include.step1 == 1]))
b1 <- paste0("At least three cognition
measures from ages 60
and above
n=", length(hrs$include.step2[hrs$include.step2 == 1]))
c1 <- paste0("Summary cogntition
status classified
n=", length(hrs$include.step3[hrs$include.step3 == 1]))
d1 <- paste0("Primary analytic
sample
n=", length(hrs$include.step10[hrs$include.step10 == 1]))
e1 <- paste0("European ancestry 
analytic sample
n=", length(hrs$include[hrs$include == "Included" & hrs$ancestry == "European"]))

# Labels for column 2
a2 <- paste0("Less than three
cognitive measures
from ages 60
and above
n=", length(hrs$include.step1[hrs$include.step2 == 0 & hrs$include.step1 == 1]))
b2 <- paste0("Cogntition status
not classified
n=", length(hrs$include.step1[hrs$include.step3 == 0 & hrs$include.step2 == 1]))
c2 <- paste0("Missing education
n=", length(hrs$include.step1[hrs$include.step4 == 0 & hrs$include.step3 == 1]), 
"\nMissing BMI
n=", length(hrs$include.step1[hrs$include.step5 == 0 & hrs$include.step4 == 1]),
"\nMissing smoking
n=", length(hrs$include.step1[hrs$include.step6 == 0 & hrs$include.step5 == 1]), 
"\nMissing 
hypertension
n=", length(hrs$include.step1[hrs$include.step7 == 0 & hrs$include.step6 == 1]),
"\nMissing diabetes
n=", length(hrs$include.step1[hrs$include.step8 == 0 & hrs$include.step7 == 1]), 
"\nMissing stroke
n=", length(hrs$include.step1[hrs$include.step9 == 0 & hrs$include.step8 == 1]),
"\nMissing alcohol
n=", length(hrs$include.step1[hrs$include.step10 == 0 & hrs$include.step9 == 1]))

d2 <- paste0("African ancestry 
analytic sample
n=", length(hrs$include[hrs$include == "Included" & hrs$ancestry == "African"]))

# Create a node dataframe
ndf <- create_node_df(
  n = 9,
  label = c(
    a1, b1, c1, d1, e1, # Column 1
    a2, b2, c2, d2
  ), # Column 2
  style = c(
    rep("solid", 5), # Column 1
    rep("dotted", 3), "solid"
  ), # Column 2
  shape = "box",
  fontname = "Arial",
  penwidth = 3,
  fixedsize = F,
  color = c(rep("black", 5), rep("silver", 3), "black")
)


# Create an edge dataframe
edf <- create_edge_df(
  from = c(
    1, 2, 3, 4, # Column 1
    1, 2, 3, 4 # Column 2
  ),
  to = c(
    2, 3, 4, 5, # Column 1
    6, 7, 8, 9 # Column 2
  ),
  arrowhead = "normal",
  style = c(rep("solid", 4), rep("dotted", 3), "solid"),
  color = c(
    rep("black", 4), # Column 1
    rep("silver", 3), "black"
  ) # Column 2
)

g <- create_graph(ndf,
  edf,
  attr_theme = NULL, directed = T
)
render_graph(g)

export_graph(g, file_name = paste0(results_dir,"HRS_PGS_Flowchart_Inclusion_Exclusion_", date, ".pdf"), file_type = "PDF")
```



## Producing reproducible descriptive statistics tables



## Compare Groups to Produce Univariate Tables
```{r univariate table}
#install.packages("compareGroups")
library(compareGroups)
table(hrs$include)
table(hrs$include.step1, hrs$include)

#PH13_KunkleOK2_PT_0.01
#PH13_KunkleFULL_PT_0.01

# Note: The official genetic variables are split by ancestry. This summary table wants them together. New variables are developed above to combine them

univar.desc.pval <- descrTable(formula = include ~ DC_60_2V.fac + DC_60_2V.miss + PH13_KunkleOK2_PT_0.01 + PH13_KunkleOK2_PT_0.01.miss+ PH13_KunkleFULL_PT_0.01 + PH13_KunkleFULL_PT_0.01.miss + APOE2010_imputed + APOE2010_imputed.miss + APOE2010_bin.fac + APOE2010_bin.miss + AGE_LAST_60 + AGE_LAST_60.miss + YEAR_LAST_60 + YEAR_LAST_60.miss + SEX.fac + SEX.miss + SCHLYRS + SCHLYRS.miss + ancestry.relevel + ancestry.miss + AHEAD_CODA + AHEAD_CODA.miss + BMI_LAST_60_v2 + BMI_LAST_60_v2.miss + HBP_EVER_LAST_60.fac + HBP_EVER_LAST_60.miss + T2D_EVER_LAST_60.fac + T2D_EVER_LAST_60.miss + EVSTROKE_LAST_60.fac + EVSTROKE_LAST_60.miss + CESD_BIN.fac + CESD_BIN.miss + SMK_CURR_LAST_60.fac + SMK_CURR_LAST_60.miss + ALC_CURR_LAST_60.fac + ALC_CURR_LAST_60.miss, data = hrs[hrs$include.step1 == 1, ], na.action = na.pass, show.all = T, show.n = T, digits.p = 3)
print(univar.desc.pval)
export2word(univar.desc.pval, file = paste0(results_dir, "HRS_PGS_Univariate_Include_P_", date, ".docx"))

univar.desc.pval <- descrTable(formula = include ~ DC_60_2V.fac + DC_60_2V.miss + PH13_KunkleOK2_PT_0.01 + PH13_KunkleOK2_PT_0.01.miss+ PH13_KunkleFULL_PT_0.01  + APOE2010_imputed+ APOE2010_bin.fac + APOE2010_bin.miss + AGE_LAST_60  + YEAR_LAST_60  + SEX.fac + SCHLYRS  + ancestry.relevel  + AHEAD_CODA  + BMI_LAST_60_v2 + BMI_LAST_60_v2.miss + HBP_EVER_LAST_60.fac  + T2D_EVER_LAST_60.fac + EVSTROKE_LAST_60.fac  + CESD_BIN.fac + SMK_CURR_LAST_60.fac  + ALC_CURR_LAST_60.fac, data = hrs[hrs$include.step1 == 1, ], na.action = na.pass, show.all = T, show.n = T, digits.p = 3, include.miss=T)
print(univar.desc.pval)
export2word(univar.desc.pval, file = paste0(results_dir, "HRS_PGS_Univariate_Include_Miss_", date, ".docx"))

```




<!-- For those using Rstudio, I've found that setting Sys.setenv('R_MAX_VSIZE'=32000000000), as has been suggested on multiple StackOverflow posts, only works on the command line, and that setting that parameter while using Rstudio does not prevent this error: -->

<!-- Error: vector memory exhausted (limit reached?) -->

<!-- After doing some more reading, I found this thread, which clarifies the problem with Rstudio, and identifies a solution, shown below: -->

<!-- Step 1: Open terminal, -->

<!-- Step 2: -->

<!-- cd ~ -->
<!-- touch .Renviron -->
<!-- open .Renviron -->

<!-- Step 3: Save the following as the first line of .Renviron: -->

<!-- R_MAX_VSIZE=100Gb  -->

<!-- Note: This limit includes both physical and virtual memory; so setting _MAX_VSIZE=16Gb on a machine with 16Gb of physical memory may not prevent this error. You may have to play with this parameter, depending on the specs of your machine -->


## Compare Groups to Produce Univariate Tables, split by ancestry
```{r}
library(compareGroups)

#PH13_KunkleOK2_PT_0.01
#PH13_KunkleFULL_PT_0.01

# Note: The official genetic variables are split by ancestry. This summary table wants them together. New variables are developed above to combine them
table(hrs$ancestry)
univar.desc.pval <- descrTable(formula = ancestry.relevel ~ DC_60_2V.fac + DC_60_2V.miss + PH13_KunkleOK2_PT_0.01 + PH13_KunkleOK2_PT_0.01.miss+ PH13_KunkleFULL_PT_0.01 + PH13_KunkleFULL_PT_0.01.miss +APOE2010_imputed + APOE2010_imputed.miss + APOE2010_bin.fac + APOE2010_bin.miss +  AGE_LAST_60 + AGE_LAST_60.miss + YEAR_LAST_60 + YEAR_LAST_60.miss + SEX.fac + SEX.miss + SCHLYRS + SCHLYRS.miss + ancestry.relevel + ancestry.miss + AHEAD_CODA + AHEAD_CODA.miss + BMI_LAST_60_v2 + BMI_LAST_60_v2.miss + HBP_EVER_LAST_60.fac + HBP_EVER_LAST_60.miss + T2D_EVER_LAST_60.fac + T2D_EVER_LAST_60.miss + EVSTROKE_LAST_60.fac + EVSTROKE_LAST_60.miss + CESD_BIN.fac + CESD_BIN.miss + SMK_CURR_LAST_60.fac + SMK_CURR_LAST_60.miss + ALC_CURR_LAST_60.fac + ALC_CURR_LAST_60.miss, data = hrs[hrs$include == "Included", ], na.action = na.pass, show.all = T, show.n = T, digits.p = 3)

print(univar.desc.pval)
export2word(univar.desc.pval, file =paste0(results_dir, "HRS_PGS_Univariate_Include_Ancestry_", date, ".docx"))

```


# Sensitivity analysis 1: Select on cohort
```{r sensitivity 1 selection}
table(include$AHEAD_CODA, useNA = "always")
sens1 <- include[include$AHEAD_CODA == "Other", ] # 8145
table(sens1$AHEAD_CODA, useNA = "always")
table(sens1$ancestry, useNA = "always")
include$sens1 <- ifelse(include$AHEAD_CODA == "Other", "Include", "Exclude")
table(include$sens1)
hrs$sens1 <- ifelse(hrs$include=="Included" & hrs$AHEAD_CODA == "Other", "Include", "Exclude")
table(hrs$sens1)

include<-hrs[hrs$include=="Included", ]

```
# Sensitivity analysis 1: Select on cohort, flow chart
```{r sensitivity 1 flow chart}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

# graph: this sets the overall layout, themes of the plot
# node: the boxes in the flowchart
# edges: the lines used to connect each of the boxes

## Need 2 columns (1, 2) with 2 rows (A-B)
# Labels for column 1
a1 <- paste0("Primary analytic sample
European n=", length(include$ancestry[include$ancestry == "European"]), "\nAfrican n=", length(include$ancestry[include$ancestry == "African"]))

b1 <- paste0("Sensitivity analysis one
 analytic sample
European n=", length(include$ancestry[include$ancestry == "European" & include$sens1 == "Include"]), "\nAfrican n=", length(include$ancestry[include$ancestry == "African" & include$sens1 == "Include"]))

# Labels for column 2
a2 <- paste0("Exclude AHEAD and
CODA cohorts
European n=", length(include$ancestry[include$ancestry == "European" & include$sens1 == "Exclude"]), "\nAfrican n=", length(include$ancestry[include$ancestry == "African" & include$sens1 == "Exclude"]))


# Create a node dataframe
ndf <- create_node_df(
  n = 3,
  label = c(
    a1, b1, # Column 1
    a2
  ), # Column 2
  style = c(
    rep("solid", 2), # Column 1
    rep("dotted", 1)
  ), # Column 2
  shape = "box",
  fontname = "Arial",
  penwidth = 3,
  fixedsize = F,
  color = c(rep("black", 2), rep("silver", 1))
)


# Create an edge dataframe
edf <- create_edge_df(
  from = c(
    1, # Column 1
    1 # Column 2
  ),
  to = c(
    2, # Column 1
    3 # Column 2
  ),
  arrowhead = "normal",
  style = c(rep("solid", 1), rep("dotted", 1)),
  color = c(
    rep("black", 1), # Column 1
    rep("silver", 1)
  ) # Column 2
)

g <- create_graph(ndf,
  edf,
  attr_theme = NULL, directed = T
)
render_graph(g)

export_graph(g, file_name =paste0(results_dir, "HRS_PGS_Flowchart_Inclusion_Exclusion_Sensitivity_One_", date, ".pdf"), file_type = "PDF")
```


# Sensitivity analysis 1: Select on cohort, inclusion table
```{r sensitivity 1 inclusion table}

#PH13_KunkleOK2_PT_0.01
#PH13_KunkleFULL_PT_0.01

univar.desc.pval <- descrTable(formula = sens1 ~ DC_60_2V.fac + DC_60_2V.miss + PH13_KunkleOK2_PT_0.01 + PH13_KunkleOK2_PT_0.01.miss+ PH13_KunkleFULL_PT_0.01 + PH13_KunkleFULL_PT_0.01.miss + APOE2010_imputed + APOE2010_imputed.miss + APOE2010_bin.fac + APOE2010_bin.miss + AGE_LAST_60 + AGE_LAST_60.miss + YEAR_LAST_60 + YEAR_LAST_60.miss + SEX.fac + SEX.miss + SCHLYRS + SCHLYRS.miss + ancestry.relevel + ancestry.miss + AHEAD_CODA + AHEAD_CODA.miss + BMI_LAST_60_v2 + BMI_LAST_60_v2.miss + HBP_EVER_LAST_60.fac + HBP_EVER_LAST_60.miss + T2D_EVER_LAST_60.fac + T2D_EVER_LAST_60.miss + EVSTROKE_LAST_60.fac + EVSTROKE_LAST_60.miss + CESD_BIN.fac + CESD_BIN.miss + SMK_CURR_LAST_60.fac + SMK_CURR_LAST_60.miss + ALC_CURR_LAST_60.fac + ALC_CURR_LAST_60.miss, data = include, na.action = na.pass, show.all = T, show.n = T, digits.p = 3)
print(univar.desc.pval)
export2word(univar.desc.pval, file = paste0(results_dir, "HRS_PGS_Univariate_Sensitivity_One_Include_", date, ".docx"))
```


# Sensitivity analysis 2: Select on health behaviors
```{r sensitivity 2 selection}

```



```{r save derived dataset with inclusion variables}
save(hrs, file=paste0(derived_data_dir,"HRS_PGS_AD_Analytic_Sample_20190814.rda"))

include<-hrs[hrs$include=="Included",]
save(include, file=paste0(derived_data_dir,"HRS_PGS_AD_Primary_Included_Sample_20190816.rda"))

ea <- hrs[hrs$ancestry == "European" & hrs$include=="Included", ]
dim(ea) # 8587

aa <- hrs[hrs$ancestry == "African" & hrs$include=="Included", ]
dim(aa) #1639

save(ea, file = paste0(derived_data_dir,"HRS_PGS_AD_Euro_Analytic_Sample_20190814.rda"))
save(aa, file = paste0(derived_data_dir,"HRS_PGS_AD_Afri_Analytic_Sample_20190814.rda"))
```

