---
title: "Alzheimer's disease PGS, APOE status, and summary cognitive status in the HRS"
subtitle: "Data QC & Variable Development"
author: "Kelly Bakulski"
date: "October 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Scripts/')
```

## Goal: Data QC and Variable Development 

```{r set source/output locations}
#Setting file paths for data, output, scripts

source_dir<-"~/Google Drive/AD_R01_WB/Data/" #location of the paxrent dataset

results_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Results/" #location of output tables
  
derived_data_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Derived_Datasets/" #location of any new datasets I'm building specifically for this paper

script_dir<-"~/Google Drive/AD_R01_WB/AD_PGS_main_effect/Scripts/" #location of this script, redundant with the knitr option above

date<-format(Sys.Date(), "%Y%m%d")
```


#Load in the dataset from sas: HRS 1995-2014
```{r load sas, eval=FALSE}
library(haven)
 hrs <- read_sas(paste0(source_dir,"hrs_democoggen_wide20190813.sas7bdat"))
  dim(hrs)
# # [1] 43232  1075 #Note, we lost 93 variables from last time (was 1172)
 hrs[hrs == ""] <- NA #There are many 
 save(hrs, file=paste0(source_dir, "hrs_democoggen_wide20190813.rda"))
```

Load in the dataset from Rda
```{r load}
#Note, can start from here without SAS step
load(paste0(source_dir, "hrs_democoggen_wide20190813.rda"))
dim(hrs)
```

##QC the dataset
#Outcome variables: Summary cogntion status
```{r outcome var}
#############################################################
#Primary outcome variable
#############################################################
table(hrs$DC_60_2V, useNA="always")
 #    1     2     3     4     5  <NA> 
 # 3866  2846  3749 12449   534 19788 
# matches expected distributions

```

#Genetics variables:APOE E4 status, PCs
```{r genetics var}

#############################################################
#APOE variables
#############################################################

table(hrs$APOE2010_bin, useNA="always") # note, not split by ancestry, use this covariate
#     0     1  <NA> 
# 11360  4205 27667 
table(hrs$APOE2010_imputed, useNA = "always") 
# e2/e2 e2/e3 e2/e4 e3/e3 e3/e4 e4/e4  <NA> 
#   111  1916   382  9333  3475   348 27667 

hrs$APOE2010_E4dose<-ifelse(hrs$APOE2010_imputed=="e2/e2" | hrs$APOE2010_imputed=="e2/e3" | hrs$APOE2010_imputed=="e3/e3", 0, NA)
hrs$APOE2010_E4dose<-ifelse(hrs$APOE2010_imputed=="e2/e4" | hrs$APOE2010_imputed=="e3/e4" , 1, hrs$APOE2010_E4dose)
hrs$APOE2010_E4dose<-ifelse(hrs$APOE2010_imputed=="e4/e4", 2, hrs$APOE2010_E4dose)
table(hrs$APOE2010_E4dose)
#111+1916+9333

table(hrs$EA_APOE4, useNA="always") #note, these frequencies of APOE4 seem really low, this is just the 2012 sample
#     0     1  <NA> 
# 11874   216 31144 
table(hrs$AA_APOE4, useNA="always") # HRS 2012 data, not for this paper
 #    0     1  <NA> 
 # 2997   103 40134 


#############################################################
#Genetic PC variables
#############################################################
summary(hrs$AA_2010_PC1)
summary(hrs$AA_2010_PC2) 
summary(hrs$AA_2010_PC3)
summary(hrs$AA_2010_PC4) 
summary(hrs$AA_2010_PC5)
summary(hrs$EA_2010_PC1)
summary(hrs$EA_2010_PC2) 
summary(hrs$EA_2010_PC3)
summary(hrs$EA_2010_PC4)  
summary(hrs$EA_2010_PC5)

# Don't see a genetic ancestry variable in the master dataset
# create one based on the PCs
table(!is.na(hrs$EA_2010_PC1))
hrs$ancestry<-ifelse(!is.na(hrs$EA_2010_PC1), "European", NA) #EA coded
hrs$ancestry<-ifelse(!is.na(hrs$AA_2010_PC1), "African", hrs$ancestry) #AA coded
table(hrs$ancestry, useNA = "always")
 # African European     <NA> 
 #    2279     9991    30962 

hrs$PC1_2010<-ifelse(!is.na(hrs$EA_2010_PC1), hrs$EA_2010_PC1, hrs$AA_2010_PC1) 
hrs$PC2_2010<-ifelse(!is.na(hrs$EA_2010_PC2), hrs$EA_2010_PC2, hrs$AA_2010_PC2) 
```
Use EAph13_IGAP_pT_1_scale and AAph13_IGAP_pT_1_scale for PGS with APOE
Pending variables for PGS without APOE
Use APOE2010_bin for APOE status for both ancestries
Use EA_2010_PC1 EA_2010_PC2 EA_2010_PC3 EA_2010_PC4 EA_2010_PC5 for European ancestry PCs
Use AA_2010_PC1 AA_2010_PC2 AA_2010_PC3 AA_2010_PC4 AA_2010_PC5 for African ancestry PCs
Make versions of these variables that keep both ancestries together for the included/excluded table



Use the new Kunkle variable: PH13_KunkleNAPOE_PT_1 for primary
PH13_KunkleYAPOE_PT_1



#Merge in Kunkle PGS variables without APOE using higher threshold
```{r Kunkle 0.01}
#############################################################
#PGS variables
#############################################################


# EAPH13_KunkleOK2_PT_0.01
# AAPH13_KunkleOK2_PT_0.01

load(paste0(derived_data_dir, "African_Kunkle_PGSs2.rda"))
dim(afro_pgs) #2279
head(afro_pgs)
str(afro_pgs)
colnames(afro_pgs)
afro_pgs$hhid<-as.character(afro_pgs$hhid)
afro_pgs$pn<-as.character(afro_pgs$pn)
afro_pgs$FID[1:10]
typeof(afro_pgs$hhid)
summary(afro_pgs$AAPH13_KunkleOK2_PT_0.01)
load(paste0(derived_data_dir,"European_Kunkle_PGSs2.rda"))
dim(euro_pgs) #9991
str(euro_pgs)
euro_pgs$hhid<-as.character(euro_pgs$hhid)
euro_pgs$pn<-as.character(euro_pgs$pn)
summary(euro_pgs$EAPH13_KunkleOK2_PT_0.01)

dim(hrs) #43232
43232-2279
hrs$HHID[1:10]
typeof(hrs$HHID)
typeof(hrs$PN)
hrs.a<-merge(hrs,afro_pgs,by.x =c("HHID", "PN"),  by.y = c("hhid", "pn"), all=T)
dim(hrs.a) #43232
summary(hrs.a$AAPH13_KunkleOK2_PT_0.01)
head(hrs.a$HHID) #40953 NA, checks out

#euro_pgs has 2 people without a corresponding hrs row, drop them
hrs.e<-merge(hrs.a,euro_pgs,by.x =c("HHID", "PN"),  by.y = c("hhid", "pn"), all.x=T, all.y=F)
dim(hrs.e) #43232
summary(hrs.e$EAPH13_KunkleOK2_PT_0.01) #33243 NA

hrs<-hrs.e

hrs$PH13_KunkleOK2_PT_0.01<-ifelse(!is.na(hrs$EAPH13_KunkleOK2_PT_0.01),hrs$EAPH13_KunkleOK2_PT_0.01, hrs$AAPH13_KunkleOK2_PT_0.01)
summary(hrs$PH13_KunkleOK2_PT_0.01) #30964 NA

#Version with APOE region
hrs$PH13_KunkleFULL_PT_0.01<-ifelse(!is.na(hrs$EAPH13_KunkleFULL_PT_0.01),hrs$EAPH13_KunkleFULL_PT_0.01, hrs$AAPH13_KunkleFULL_PT_0.01 )
summary(hrs$PH13_KunkleFULL_PT_0.01)
```
hrs$PH13_KunkleOK2_PT_0.01 is the primary variable fo interest


#Primary model variables (age at last visit, year of last visit, sex, years of education, genetic ancestry)
```{r key covariates}
#############################################################
#Key covariates
#############################################################
summary(hrs$AGE_LAST_60) #age at last visit, no values before age 60
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NAs 
  # 60.00   67.00   75.00   75.34   83.00  109.00   15792 

summary(hrs$YEAR_LAST_60) #year of last visit
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   # 1995    2004    2012    2009    2014    2014   15792 


table(hrs$GENDER, useNA = "always")
#     1     2  <NA> 
# 19081 24129    22
colnames(hrs)[colnames(hrs)=="GENDER"]<-"SEX" #rename GENDER as SEX appropriately
table(hrs$SEX, useNA = "always")

summary(hrs$SCHLYRS) #Erin updated 99's to missing in this in the datastet 20190813
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   # 0.00   11.00   12.00   12.13   14.00   17.00    1240 

```
Use HRS variables: AGE_LAST_60, YEAR_LAST_60, GENDER
Use new variables: SCHLYRS_miss, ancestry



# Sensitivity model variables (cohort, BMI at last visit, ever hypertension, ever diabetes, smoking status, depression status, ever stroke)
```{r sensitivity covariates}
#############################################################
#Sensitivity model covariates
#############################################################

table(hrs$STUDY, useNA = "always") #USE this instead of wtcohort to capture the spouses correctly. 
# Leave 11 (AHEAD) and 21 (CODA)
hrs$AHEAD_CODA<-ifelse(hrs$STUDY==11, "AHEAD", NA) #1914-1923, AHEAD
hrs$AHEAD_CODA<-ifelse(hrs$STUDY==21, "CODA", hrs$AHEAD_CODA) #1924-1930, CODA
hrs$AHEAD_CODA<-ifelse(hrs$STUDY==1 | hrs$STUDY>30, "Other", hrs$AHEAD_CODA)
table(hrs$AHEAD_CODA, useNA="always")


# BMI
summary(hrs$BMI_LAST_60) #Red flag, we have a ton of missing, Erin working on this
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  # 11.50   24.85   28.20   29.19   32.51   75.80   24748 
hrs$BMI_LAST_60_v2<-hrs$BMI_LAST_60
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60) & hrs$YEAR_LAST_60==2006, hrs$BMI_H04, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60_v2) & hrs$YEAR_LAST_60==2006, hrs$R7BMI, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60) & hrs$YEAR_LAST_60==2008, hrs$BMI_H06, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60_v2) & hrs$YEAR_LAST_60==2008, hrs$R8BMI, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60) & hrs$YEAR_LAST_60==2010, hrs$BMI_H08, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60_v2) & hrs$YEAR_LAST_60==2010, hrs$R9BMI, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60) & hrs$YEAR_LAST_60==2012, hrs$BMI_H10, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60_v2) & hrs$YEAR_LAST_60==2012, hrs$R10BMI, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60) & hrs$YEAR_LAST_60==2014, hrs$BMI_H12, hrs$BMI_LAST_60_v2)
hrs$BMI_LAST_60_v2<-ifelse(is.na(hrs$BMI_LAST_60_v2) & hrs$YEAR_LAST_60==2014, hrs$R11BMI, hrs$BMI_LAST_60_v2)
summary(hrs$BMI_LAST_60_v2)


table(hrs$EVSTROKE_LAST_60, useNA = "always")
#     0     1  <NA> 
# 22586  3684 16962 
hrs$EVSTROKE_LAST_60.fac<-factor(hrs$EVSTROKE_LAST_60, levels=c(0,1), labels=c("No", "Yes"))
table(hrs$EVSTROKE_LAST_60.fac, useNA = "always")

#develop from Erin's code
# SMK_CURR_LAST_60
# ALC_CURR_LAST_60
# HBP_EVER_LAST_60
# T2D_EVER_LAST_60
# CESD_AVG #After age 60

#CODE TO CREATE alcohol, smoking (ever, current, never), hypertension, diabetes, cesd_avg)

# Alcohol
table(hrs$YEAR_LAST_60[hrs$YEAR_LAST_60==2006])
table(hrs$R8DRINK, useNA = "always")
hrs$ALC_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2006, hrs$R8DRINK, NA)
hrs$ALC_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2008, hrs$R9DRINK, hrs$ALC_CURR_LAST_60)
hrs$ALC_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2010, hrs$R10DRINK, hrs$ALC_CURR_LAST_60)
hrs$ALC_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2012, hrs$R11DRINK, hrs$ALC_CURR_LAST_60)
hrs$ALC_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2014, hrs$R12DRINK, hrs$ALC_CURR_LAST_60)
table(hrs$ALC_CURR_LAST_60, useNA="always")
hrs$ALC_CURR_LAST_60.fac<-factor(hrs$ALC_CURR_LAST_60, levels=c(0,1), labels=c("No", "Yes"))
table(hrs$ALC_CURR_LAST_60.fac, useNA="always")

# Smoking
table(hrs$smoke06, useNA = "always")
hrs$SMK_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2006, hrs$smoke06, NA)
hrs$SMK_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2008, hrs$smoke08, hrs$SMK_CURR_LAST_60)
hrs$SMK_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2010, hrs$smoke10, hrs$SMK_CURR_LAST_60)
hrs$SMK_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2012, hrs$smoke12, hrs$SMK_CURR_LAST_60)
hrs$SMK_CURR_LAST_60<-ifelse(hrs$YEAR_LAST_60==2014, hrs$smoke14, hrs$SMK_CURR_LAST_60)
table(hrs$SMK_CURR_LAST_60, useNA="always")
hrs$SMK_CURR_LAST_60.fac<-factor(hrs$SMK_CURR_LAST_60)
table(hrs$SMK_CURR_LAST_60.fac, useNA="always")

# Hypertension
table(hrs$R8HIBPE, useNA = "always")
hrs$HBP_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2006, hrs$R8HIBPE, NA)
hrs$HBP_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2008, hrs$R9HIBPE, hrs$HBP_EVER_LAST_60)
hrs$HBP_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2010, hrs$R10HIBPE, hrs$HBP_EVER_LAST_60)
hrs$HBP_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2012, hrs$R11HIBPE, hrs$HBP_EVER_LAST_60)
hrs$HBP_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2014, hrs$R12HIBPE, hrs$HBP_EVER_LAST_60)
table(hrs$HBP_EVER_LAST_60, useNA="always")
hrs$HBP_EVER_LAST_60.fac<-factor(hrs$HBP_EVER_LAST_60, levels=c(0,1), labels=c("No", "Yes"))
table(hrs$HBP_EVER_LAST_60.fac, useNA="always")

# Diabetes
table(hrs$R8DIABE, useNA = "always")
hrs$T2D_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2006, hrs$R8HIBPE, NA)
hrs$T2D_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2008, hrs$R9DIABE, hrs$T2D_EVER_LAST_60)
hrs$T2D_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2010, hrs$R10DIABE, hrs$T2D_EVER_LAST_60)
hrs$T2D_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2012, hrs$R11DIABE, hrs$T2D_EVER_LAST_60)
hrs$T2D_EVER_LAST_60<-ifelse(hrs$YEAR_LAST_60==2014, hrs$R12DIABE, hrs$T2D_EVER_LAST_60)
table(hrs$T2D_EVER_LAST_60, useNA="always")
hrs$T2D_EVER_LAST_60.fac<-factor(hrs$T2D_EVER_LAST_60, levels=c(0,1), labels=c("No", "Yes"))
table(hrs$T2D_EVER_LAST_60.fac, useNA="always")

#Depression
summary(hrs$R3CESD)
hrs$CESD_AVG<-ifelse(hrs$YEAR_LAST_60==2006, rowMeans(hrs[,c("R3CESD", "R4CESD","R5CESD", "R6CESD", "R7CESD", "R8CESD")], na.rm = T), NA)
hrs$CESD_AVG<-ifelse(hrs$YEAR_LAST_60==2008, rowMeans(hrs[,c("R3CESD", "R4CESD","R5CESD", "R6CESD", "R7CESD", "R8CESD", "R9CESD")], na.rm = T), hrs$CESD_AVG)
hrs$CESD_AVG<-ifelse(hrs$YEAR_LAST_60==2010, rowMeans(hrs[,c("R3CESD", "R4CESD","R5CESD", "R6CESD", "R7CESD", "R8CESD", "R9CESD", "R10CESD")], na.rm = T), hrs$CESD_AVG)
hrs$CESD_AVG<-ifelse(hrs$YEAR_LAST_60==2012, rowMeans(hrs[,c("R3CESD", "R4CESD","R5CESD", "R6CESD", "R7CESD", "R8CESD", "R9CESD", "R10CESD", "R11CESD")], na.rm = T), hrs$CESD_AVG)
hrs$CESD_AVG<-ifelse(hrs$YEAR_LAST_60==2014, rowMeans(hrs[,c("R3CESD", "R4CESD","R5CESD", "R6CESD", "R7CESD", "R8CESD", "R9CESD", "R10CESD", "R11CESD", "R12CESD")], na.rm = T), hrs$CESD_AVG)
summary(hrs$CESD_AVG)

length(hrs$CESD_AVG[hrs$ancestry=="African" & hrs$CESD_AVG>=1.8295108])
table(hrs$ancestry=="African", hrs$CESD_AVG>=1.8295108, useNA = "always") #Expect 688
hrs$CESD_BIN<-ifelse(hrs$ancestry=="African" & hrs$CESD_AVG>=1.8295108, 1, NA)
hrs$CESD_BIN<-ifelse(hrs$ancestry=="African" & hrs$CESD_AVG<1.8295108, 0, hrs$CESD_BIN)
table(hrs$CESD_BIN, useNA = "always") # checks out

table(hrs$ancestry=="European", hrs$CESD_AVG>=1.20643, useNA = "always") #Expect 3137
hrs$CESD_BIN<-ifelse(hrs$ancestry=="European" & hrs$CESD_AVG>=1.20643, 1, hrs$CESD_BIN)
hrs$CESD_BIN<-ifelse(hrs$ancestry=="European" & hrs$CESD_AVG<1.20643, 0, hrs$CESD_BIN)
table(hrs$CESD_BIN, useNA = "always")
hrs$CESD_BIN.fac<-factor(hrs$CESD_BIN, levels=c(0,1), labels=c("No", "Yes"))
table(hrs$CESD_BIN.fac, useNA="always")

```
Use parent dataset variables: EVSTROKE_LAST_60
Use new variables: AHEAD_CODA, SMK_CURR_LAST_60, ALC_CURR_LAST_60, HBP_EVER_LAST_60, T2D_EVER_LAST_60, CESD_BIN, BMI_LAST_V2 #After age 60 


#Save Rda with newly derived variables (AHEAD_CODA, stroke_conserv, SCHLYRS_nomiss, ancestry)
```{r save derived dataset}

save(hrs,  file=paste0(derived_data_dir,"hrs_democoggen_wide20190813.rda"))

```

